---
output: github_document
always_allow_html: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

library(kableExtra)
```

# MachineShop: Machine Learning Models and Tools

## Overview

`MachineShop` is a meta-package for statistical and machine learning with a common interface for model fitting, prediction, performance assessment, and presentation of results.  Support is provided for predictive modeling of numerical, categorical, and censored time-to-event outcomes, including those listed in the table below, and for resample (bootstrap and cross-validation) estimation of model performance.

<div>
```{r echo = FALSE}
library(MachineShop)

modelnames <- c("C5.0 Classification" = "C50Model",
                "Conditional Inference Trees" = "CForestModel",
                "Cox Regression" = "CoxModel",
                "Generalized Linear Models" = "GLMModel",
                "Gradient Boosted Models" = "GBMModel",
                "Lasso and Elastic-Net" = "GLMNetModel",
                "Feed-Forward Neural Networks" = "NNetModel",
                "Partial Least Squares" = "PLSModel",
                "Ordered Logistic Regression" = "POLRModel",
                "Random Forests" = "RandomForestModel",
                "Stacked Regression" = "StackedModel",
                "Super Learner" = "SuperModel",
                "Parametric Survival Regression" = "SurvRegModel",
                "Support Vector Machines" = "SVMModel",
                "Extreme Gradient Boosting" = "XGBModel")

types <- c("factor", "numeric", "ordered", "Surv")
x <- lapply(modelnames, function(modelname) {
  model <- get(modelname)()
  structure(c(modelname, ifelse(types %in% model@types, "x", "")),
            names = c("Constructor", types))
})
df <- as.data.frame(do.call(rbind, x))

kable(df, align = "c") %>%
  kable_styling("striped", full_width = FALSE, position = "center") %>%
  add_header_above(c(" " = 1, " " = 1, "Response Variable Types" = 4))
```
</div>

## Installation

```{r eval = FALSE}
# Current release from CRAN
install.packages("MachineShop")

# Development version from GitHub
# install.packages("devtools")
devtools::install_github("brian-j-smith/MachineShop")

# Development version with vignettes
devtools::install_github("brian-j-smith/MachineShop", build_vignettes = TRUE)
```


## Example

The following is a brief example using the package to apply gradient boosted models to predict the species of flowers in Edgar Anderson's iris data set.

### Training and Test Set Analysis

```{r}
## Load the package
library(MachineShop)
library(magrittr)

## Iris flower species (3 level response) data set
df <- iris
df$Species <- factor(df$Species)

## Training and test sets
set.seed(123)
trainindices <- sample(nrow(df), nrow(df) * 2 / 3)
train <- df[trainindices, ]
test <- df[-trainindices, ]

## Model formula
fo <- Species ~ .

## Gradient boosted mode fit to training set
gbmfit <- fit(fo, data = train, model = GBMModel)

## Variable importance
(vi <- varimp(gbmfit))

plot(vi)
```

```{r}
## Test set predicted probabilities
predict(gbmfit, newdata = test, type = "prob") %>% head

## Test set predicted classifications
predict(gbmfit, newdata = test) %>% head

## Test set performance
obs <- response(fo, data = test)
pred <- predict(gbmfit, newdata = test, type = "prob")
modelmetrics(obs, pred)
```

### Resampling

```{r, fig.height = 4}
## Resample estimation of model performance
(perf <- resample(fo, data = df, model = GBMModel, control = CVControl))

summary(perf)

plot(perf)
```

### Model Tuning

```{r, fig.height = 4}
## Model tuning
gbmtune <- tune(fo, data = df, model = GBMModel,
                grid = expand.grid(n.trees = c(25, 50, 100),
                                   interaction.depth = 1:3,
                                   n.minobsinnode = c(5, 10)))

plot(gbmtune, type = "line")
```

```{r}
## Fit the tuned model
gbmtunefit <- fit(fo, data = df, model = gbmtune)

varimp(gbmtunefit)
```

### Model Comparisons

```{r}
## Model comparisons
control <- CVControl(folds = 10, repeats = 5)

gbmperf <- resample(fo, data = df, model = GBMModel, control = control)
rfperf <- resample(fo, data = df, model = RandomForestModel, control = control)
nnetperf <- resample(fo, data = df, model = NNetModel, control = control)

perf <- Resamples(GBM = gbmperf, RF = rfperf, NNet = nnetperf)

summary(perf)

plot(perf)
```

```{r}
## Pairwise model differences and t-tests
perfdiff <- diff(perf)

summary(perfdiff)

t.test(perfdiff)

plot(perfdiff)
```

### Recipes

```{r}
library(recipes)

rec <- recipe(fo, data = df) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors()) %>%
  step_pca(all_predictors())

perf <- resample(rec, model = GBMModel, control = CVControl)

summary(perf)
```


## Documentation

Once the package is installed, general documentation on its usage can be viewed with the following console commands.

```{r eval = FALSE, message = FALSE}
library(MachineShop)

# Package help summary
?MachineShop

# Vignette
RShowDoc("Introduction", package = "MachineShop")
```

