---
output: github_document
always_allow_html: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

library(kableExtra)
```

# MachineShop: Machine Learning Models and Tools

## Overview

`MachineShop` is a meta-package for statistical and machine learning with a common interface for model fitting, prediction, performance assessment, and presentation of results.  Support is provided for predictive modeling of numerical, categorical, and censored time-to-event outcomes, including those listed in the table below, and for resample (bootstrap and cross-validation) estimation of model performance.

<div>
```{r echo = FALSE}
library(MachineShop)

df <- data.frame(factor = character(), numeric = character(), ordered = character(), Surv = character(), stringsAsFactors = FALSE)

models <- c("C5.0 Classification" = C50Model,
            "Conditional Inference Trees" = CForestModel,
            "Cox Regression" = CoxModel,
            "Generalized Linear Models" = GLMModel,
            "Gradient Boosted Models" = GBMModel,
            "Lasso and Elastic-Net" = GLMNetModel,
            "Feed-Forward Neural Networks" = NNetModel,
            "Partial Least Squares" = PLSModel,
            "Ordered Logistic Regression" = POLRModel,
            "Random Forests" = RandomForestModel,
            "Parametric Survival Regression" = SurvRegModel,
            "Support Vector Machines" = SVMModel)

for(modelname in names(models)) {
  model <- models[[modelname]]()
  df[modelname,] <- ifelse(names(df) %in% model@types, "x", "")
}

kable(df, align = "c") %>%
  kable_styling("striped", full_width = FALSE, position = "center") %>%
  add_header_above(c(" " = 1, "Response Variable Types" = 4))
```
</div>

## Installation

```{r eval = FALSE}
# Development version from GitHub
# install.packages("devtools")
devtools::install_github("brian-j-smith/MachineShop")

# Development version with vignettes
devtools::install_github("brian-j-smith/MachineShop", build_vignettes = TRUE)
```


## Example

The following is a brief example using the package to apply gradient boosted models to predict the species of flowers in Edgar Anderson's iris dataset.

```{r}
## Load the package
library(MachineShop)
library(magrittr)

## Iris flower species (3 level response)
df <- iris
df$Species <- factor(df$Species)

## Create training and test sets
set.seed(123)
trainindices <- sample(nrow(df), nrow(df) * 2 / 3)
train <- df[trainindices, ]
test <- df[-trainindices, ]

## Gradient boosted mode fit to training set
gbmfit <- fit(GBMModel(), Species ~ ., data = train)

## Variable importance
(vi <- varimp(gbmfit))

plot(vi)
```

```{r}
## Test set predicted probabilities
predict(gbmfit, newdata = test, type = "prob") %>% head

## Test set predicted classification
predict(gbmfit, newdata = test) %>% head
```

```{r, fig.height = 4}
## Resample estimation of model performance
(perf <- resample(GBMModel(), Species ~ ., data = df))

summary(perf)

plot(perf, metrics = c("Accuracy", "Kappa", "MLogLoss"))
```

```{r, fig.height = 4}
## Model tuning
gbmtune <- tune(GBMModel, Species ~ ., data = df,
                grid = expand.grid(n.trees = c(25, 50, 100),
                                   interaction.depth = 1:3,
                                   n.minobsinnode = c(5, 10)))

plot(gbmtune, type = "line", metrics = c("Accuracy", "Kappa", "MLogLoss"))
```

## Documentation

Once the package is installed, general documentation on its usage can be viewed with the following console commands.

```{r eval = FALSE, message = FALSE}
library(MachineShop)

# Package help summary
?MachineShop

# Vignette
RShowDoc("Introduction", package = "MachineShop")
```

