# Model Information

A named list of package-supplied models can be obtained interactively with the `modelinfo` function, and includes a descriptive `"label"` for each, source `"packages"` on which the models depend, supported response variable `"types"`, and `"arguments"` that can be specified in calls to the model functions.  Function `modelinfo` can be called without arguments, with one or more model functions, variable types, or response variables; and will return information on all models matching the calling arguments.

```{r ex_modelinfo}
## Analysis libraries
library(MachineShop)
library(magrittr)

## All availble models
modelinfo() %>% names

## Model-specific information
modelinfo(C50Model, CoxModel)
```


## Type-Specific

```{r ex_modelinfo_type}
## All factor response-specific models
modelinfo(factor(0)) %>% names

## Identify factor response-specific models
modelinfo(factor(0), AdaBagModel, C50Model, CoxModel) %>% names
```


## Response Variable-Specific

```{r ex_modelinfo_response}
## Models for a responses variable
modelinfo(iris$Species) %>% names
```


# Metric Information

A named list of supplied metrics can be obtained with the `metricinfo` function, and includes a descriptive `"label"` for each, whether to `"maximize"` the metrics for better performance, their function `"arguments"`, and supported observed and predicted response variable `"types"`.  Function `metricinfo` may be called without arguments, with one or more metric functions, an observed response variable, an observed and predicted response variable pair, response variable types, or resampled output; and will return information on all matching metrics.

```{r ex_metricinfo}
## Analysis libraries
library(MachineShop)
library(magrittr)

## All availble metrics
metricinfo() %>% names

## Metric-specific information
metricinfo(auc, r2)
```


## Type-Specific

```{r ex_metricinfo_type}
## Metrics for observed and predicted response variable types
metricinfo(factor(0)) %>% names

metricinfo(factor(0), factor(0)) %>% names

metricinfo(factor(0), matrix(0)) %>% names

## Identify factor-specific metrics
metricinfo(factor(0), accuracy, auc, r2) %>% names
```


## Response Variable-Specific

```{r ex_metricinfo_response}
## Metrics for observed and predicted responses from a model fit
model_fit <- fit(Species ~ ., data = iris, model = C50Model)
obs <- response(model_fit)
pred <- predict(model_fit, type = "prob")
metricinfo(obs, pred) %>% names
```


## Resample-Specific

```{r ex_metricinfo_resample}
## Metrics for resampled output
model_res <- resample(Species ~ ., data = iris, model = C50Model)
metricinfo(model_res) %>% names
```
