# Description

**MachineShop** is a meta-package for statistical and machine learning with a unified interface for model fitting, prediction, performance assessment, and presentation of results.  Support is provided for predictive modeling of numerical, categorical, and censored time-to-event outcomes and for resample (bootstrap, cross-validation, and split training-test sets) estimation of model performance.  This vignette introduces the package interface with a survival data analysis example, followed by supported methods of variable specification; applications to other response variable types; available performance metrics, resampling techniques, and graphical and tabular summaries; and modeling strategies.


# Features

```{r overview_modelinfo, echo=FALSE}
library(MachineShop)
info <- modelinfo()
```

* Unified and concise interface for model fitting, prediction, and performance assessment.
* Current support for `r length(info)` established models from `r length(unique(sapply(info, function(x) x$packages[1])))` **R** packages.
* Dynamic model parameters.
* Ensemble modeling with stacked regression and super learners.
* Modeling of response variables types: binary factors, multi-class nominal and ordinal factors, numeric vectors and matrices, and censored time-to-event survival.
* Model specification with traditional formulas, design matrices, and flexible pre-processing [recipes](https://cran.r-project.org/package=recipes).
* Resample estimation of predictive performance, including cross-validation, bootstrap resampling, and split training-test set validation.
* Parallel execution of resampling algorithms.
* Choices of performance metrics: accuracy, areas under ROC and precision recall curves, Brier score, coefficient of determination (R^2^), concordance index, cross entropy, F score, Gini coefficient, unweighted and weighted Cohen's kappa, mean absolute error, mean squared error, mean squared log error, positive and negative predictive values, precision and recall, and sensitivity and specificity.
* Graphical and tabular performance summaries: calibration curves, confusion matrices, partial dependence plots, performance curves, lift curves, and variable importance.
* Model tuning over automatically generated grids of parameter values and randomly sampled grid points.
* Model selection and comparisons for any combination of models and model parameter values.
* User-definable models and performance metrics.


# Getting Started


## Installation

```{r overview_install, eval = FALSE}
# Current release from CRAN
install.packages("MachineShop")

# Development version from GitHub
# install.packages("devtools")
devtools::install_github("brian-j-smith/MachineShop")

# Development version with vignettes
devtools::install_github("brian-j-smith/MachineShop", build_vignettes = TRUE)
```


## Documentation

Once installed, the following **R** commands will load the package and display its help system documentation.  Online documentation and examples are available at the [MachineShop website](https://brian-j-smith.github.io/MachineShop/).

```{r overview_docs, eval = FALSE, message = FALSE}
library(MachineShop)

# Package help summary
?MachineShop

# Vignette
RShowDoc("Introduction", package = "MachineShop")
```
