# Response Variable Types

The **R** class types of response variables play a central role in their analysis with the package.  They determine, for example, the specific models that can be fit, fitting algorithms employed, predicted values produced, and applicable performance metrics and analyses.  As described in the following sections, factors, ordered factors, numeric vectors and matrices, and survival objects are supported by the package.


## Factor

Categorical responses with two or more levels should be coded as a `factor` variable for analysis.  Prediction is of factor levels by default and of level-specific probabilities if `type = "prob"`.

```{r using_responses_factor}
## Iris flowers species (3-level factor)
model_fit <- fit(Species ~ ., data = iris, model = GBMModel)
predict(model_fit) %>% head
predict(model_fit, type = "prob") %>% head
```

In the case of a binary factor, the second factor level is treated as the event and the level for which predicted probabilities are computed. 

```{r using_responses_factor_binary}
## Pima Indians diabetes statuses (binary factor)
library(MASS)

model_fit <- fit(type ~ ., data = Pima.tr, model = GBMModel)
predict(model_fit, newdata = Pima.te) %>% head
predict(model_fit, newdata = Pima.te, type = "prob") %>% head
```


## Ordered Factor

Categorical responses can be designated as having ordered levels by storing them as an `ordered` factor variable.  For categorical vectors, this can be accomplished with the `factor()` function and its argument `ordered = TRUE` or more simply with the `ordered()` function.  Numeric vectors can be converted to ordered factors with the `cut()` function.

```{r using_responses_ordered}
## Iowa City housing prices (ordered factor)
df <- within(ICHomes,
  sale_amount <- cut(sale_amount, breaks = 3,
                     labels = c("Low", "Medium", "High"),
                     ordered_result = TRUE)
)

model_fit <- fit(sale_amount ~ ., data = df, model = GBMModel)
predict(model_fit) %>% head
predict(model_fit, type = "prob") %>% head
```


## Numeric Vector

Code univariate numerical responses as a `numeric` variable.  Predicted numeric values are of the original storage mode (integer or double) by default, and doubles if `type = "prob"`.

```{r using_responses_numeric}
## Iowa City housing prices
model_fit <- fit(sale_amount ~ ., data = ICHomes, model = GBMModel)
predict(model_fit) %>% head
predict(model_fit, type = "prob") %>% head
```


## Numeric Matrix

Store multivariate numerical responses as a numeric `matrix` variable for model fitting with traditional formulas and model frames.

```{r using_responses_matrix}
## Anscombe's multiple regression models dataset

## Numeric matrix response formula
model_fit <- fit(cbind(y1, y2, y3) ~ x1, data = anscombe, model = LMModel)
predict(model_fit) %>% head
```

For recipes, the multiple response may be defined as a single variable in a data frame or on the left hand side of a recipe formula.

```{r using_responses_matrix_recipe}
## Numeric matrix response recipe
## Defined in a data frame
df <- within(anscombe, y <- cbind(y1, y2, y3))
rec <- recipe(y ~ x1, data = df)

## Defined in a recipe formula
rec <- recipe(y1 + y2 + y3 ~ x1, data = anscombe)
model_fit <- fit(rec, model = LMModel)
predict(model_fit) %>% head
```


## Survival Objects

Censored time-to-event survival responses should be stored as a `Surv` variable for model fitting with traditional formulas and model frames.

```{r using_responses_surv, results="hide"}
## Survival response formula
library(survival)

fit(Surv(time, status) ~ ., data = surv_train, model = GBMModel)
```

For recipes, survival responses may be defined as a single `Surv` variable in a data frame or with the individual survival time and event variables given on the left hand side of a recipe formula and their roles designated with the `role_surv()` function.

```{r using_responses_surv_recipe, results="hide"}
## Survival response recipe
## Defined in a data frame
df <- within(veteran, {
  y <- Surv(time, status)
  remove(time, status)
})
rec <- recipe(y ~ ., data = df)

## Defined in a recipe formula
rec <- recipe(time + status ~ ., data = veteran) %>%
  role_surv(time = time, event = status)
fit(rec, model = GBMModel)
```
