# Response Variable Types

The R class types of response variables play a central role in their analysis with the package.  They determine, for example, the specific models that can be fit, fitting algorithms employed, predicted values produced, and applicable performance metrics and analyses.  As described below, factors, ordered factors, numeric vectors and matrices, and survival responses are supported by the package.


## Factors

Categorical responses with two or more levels should be coded as `factor` variables for analysis.

```{r results="hide"}
## Iris flowers species (3-level factor)
fit(Species ~ ., data = iris, model = GBMModel)
```

```{r results="hide"}
## Pima Indians diabetes statuses (binary factor)
library(MASS)

fit(type ~ ., data = Pima.tr, model = GBMModel)
```


## Ordered Factors

Ordinal categorical responses should be coded as `ordered` factor variables.  For categorical vectors, this can be accomplished  with the `factor` function and its argument `ordered = TRUE` or more simply with the `ordered` function.  Numeric vectors can be converted to ordered factors with the `cut` function.

```{r results="hide"}
## Boston housing prices (ordered factor)
library(MASS)

df <- within(Boston, {
  medv <- cut(medv, breaks = 3, ordered_result = TRUE)
})

fit(medv ~ ., data = df, model = GBMModel)
```


## Numeric Vectors

Univariate numerical responses should be coded as `numeric` variables.

```{r results="hide"}
## Boston housing prices
library(MASS)

fit(medv ~ ., data = Boston, model = GBMModel)
```


## Numeric Matrices

Multivariate numerical responses should be given as numeric `matrix` variables for model fitting with traditional formulas or model frames.

```{r results="hide"}
## Anscombe's multiple regression models dataset

## Numeric matrix response formula
fit(cbind(y1, y2, y3) ~ x1, data = anscombe, model = LMModel)
```

For recipes, the multiple response variables should be given on the left hand side of the formula specification.

```{r results="hide"}
## Numeric matrix response recipe
rec <- recipe(y1 + y2 + y3 ~ x1, data = anscombe)

fit(rec, model = LMModel)
```


## Survival Objects

Survival responses should be coded as `Surv` variables for model fitting with traditional formulas or model frames.

```{r results="hide"}
## Survival response formula
library(survival)

fit(Surv(time, status) ~ ., data = surv_df, model = GBMModel)
```

For recipes, survival outcomes should be specified with the individual survival time and event variables given on the left hand side of the formula and with their roles designated as `"surv_time"` and `"surv_event"`.

```{r results="hide"}
## Survival response recipe
rec <- recipe(time + status ~ ., data = surv_df) %>%
  add_role(time, new_role = "surv_time") %>%
  add_role(status, new_role = "surv_event")

fit(rec, model = GBMModel)
```
