# Model Fit and Prediction

Model fitting requires user specification of an available model.  A named list of `MachineShop` models can be obtained interactively with the `modelinfo` function, and includes a descriptive `"label"`, the source `"packages"` on which the models depend, supported response variable `"types"`, and `"arguments"` that can be specified in calls to the model functions.  Note that in order to use a model the source packages must be installed with the `install.packages` or equivalent function, but need not be loaded with the `library` function.  Function `modelinfo` can be called with one or more model functions, function names, function calls, or observed response variables; and will return information on all models matching the calling arguments.

```{r}
## All available models
modelinfo() %>% names

## Survival-specific models
modelinfo(Surv(0)) %>% names

## Model-specific information
modelinfo(GBMModel)
```

Information is displayed above for the `GBMModel` function which is a generalized boosted regression model --- a tree-based ensemble method that can be applied to survival outcomes.  Package models, like `GBMModel` can be specified in the `model` argument of the `fit` function to estimate a relationship (`surv_fo`) between predictors and an outcome based on a set of data (`surv_train`).  Argument specifications may be in terms of the model function, function name, or a function call.

```{r results="hide"}
## Generalized boosted regression fit

## Model function
surv_fit <- fit(surv_fo, data = surv_train, model = GBMModel)

## Model function name
fit(surv_fo, data = surv_train, model = "GBMModel")

## Model function call
fit(surv_fo, data = surv_train, model = GBMModel(n.trees = 100, interaction.depth = 1))
```

A `predict` function is supplied for application to model fit results to obtain predicted values on a dataset specified with its `newdata` argument or on the original dataset if not specified.  Survival means are predicted for survival outcomes by default.  Estimates of the associated survival distributions are needed to calculate the means.  For models, like `GBMModel`, that perform semi- or non-parametric survival analysis, Weibull approximations to the survival distributions are the default for mean estimation.  Other choices of distributional approximations are exponential, Rayleigh, and empirical.  Empirical distributions are applicable to Cox proportional hazards-based models and can be calculated with the method of Breslow [-@breslow:1972:DPC], Efron [-@efron:1977:ECL, default], or Fleming and Harrington [-@fleming:1984:NES].  Note, however, that empirical survival means are undefined mathematically if an event does not occur at the longest follow-up time.  In such situations, a restricted survival mean is calculated by changing the longest follow-up time to an event, as suggested by Efron [-@efron:1967:PFB], which will be negatively biased.  

```{r}
## Predicted survival means (default: Weibull distribution)
predict(surv_fit, newdata = surv_test) %>% head

## Predicted survival means (empirical distribution)
predict(surv_fit, newdata = surv_test, dist = "empirical") %>% head
```

In addition to survival means, predicted survival probabilities (`type = "prob"`) or 0-1 survival events (default: `type = "response"`) can be obtained with the follow-up `times` argument.  The cutoff probability for classification of survival events (or other binary responses) can be set optionally with the `cutoff` argument (default: `cutoff = 0.5`).  As before, distributional approximations to the survival functions may be specified for the predictions, with the default for survival probabilities being the empirical distribution.

```{r}
## Predict survival probabilities and events at specified follow-up times
surv_times <- 365 * c(5, 10)

predict(surv_fit, newdata = surv_test, times = surv_times, type = "prob") %>% head

predict(surv_fit, newdata = surv_test, times = surv_times, cutoff = 0.5) %>% head
```
