# Package Extensions

Custom models and metrics can be defined with `MLModel` and `MLMetric` for use with the model fitting, prediction, and performance assessment tools provided by the package.


## MLModel

The `MLModel` function creates a model object that can be used with the previously described fitting functions.  It take the following arguments.

`name`
  : Character name of the object to which the model is assigned.
  
`label`
  : Optional character descriptor for the model (default: `name`).
  
`packages`
  : Character vector of source packages required to use the model.
  
`types`
  : Character vector of response variable types to which the model can be fit.  Supported types are `"binary"`, `"factor"`, `"matrix"`, `"numeric"`, `"ordered"`, and `"Surv"`.
  
`fit`
  : Model fitting function whose arguments are a `formula`, a `ModelFrame` named `data`, case `weights`, and an ellipsis.  Argument `data` may be converted to a data frame with the `as.data.frame` function as needed.  The fit function should return the object resulting from the model fit.
  
`predict`
  : Prediction function whose arguments are the `object` returned by `fit`, a `ModelFrame` named
`newdata` of predictor variables, optional vector of `times` at which to predict survival, and an ellipsis.  Argument `data` may be converted to a data frame with the `as.data.frame` function as needed.  Values returned by the function should be formatted according to the response variable types below.

* `factor` : Vector or column matrix of probabilities for the second level of binary factors or a matrix whose columns contain the probabilities for factors with more than two levels.
* `matrix` : Matrix of predicted responses.
* `numeric` : Vector or column matrix of predicted responses.
* `Surv` : Matrix whose columns contain survival probabilities at `times` if supplied or a vector of predicted survival means otherwise.
  
`varimp`
  : Variable importance function whose arguments are the `object` returned by `fit`, optional arguments passed from calls to `varimp`, and an ellipsis.  The function should return a vector of importance values named after the predictor variables or a matrix or data frame whose rows are named after the predictors.

```{r using_extensions_mlmodel}
## Logistic regression model extension
LogisticModel <- MLModel(
  name = "LogisticModel",
  label = "Logistic Model",
  types = "binary",
  fit = function(formula, data, weights, ...) {
    glm(formula, data = as.data.frame(data), weights = weights,
        family = binomial, ...)
  },
  predict = function(object, newdata, ...) {
    predict(object, newdata = as.data.frame(newdata), type = "response")
  },
  varimp = function(object, ...) {
    pchisq(coef(object)^2 / diag(vcov(object)), 1)
  }
)
```


## MLMetric

The `MLMetric` function creates a metric object that can be used as previously described for the model performance metrics.  Its first argument is a function to compute the metric, defined to accept `observed` and `predicted` as the first two arguments and with an ellipsis to accommodate others.  Its remaining arguments are as follows.

`name`
  :  Character name of the object to which the metric is assigned.
  
`label`
  : Optional character descriptor for the metric (default: `name`).
  
`maximize`
  : Logical indicating whether higher values of the metric correspond to better predictive performance.

```{r using_extensions_mlmetric}
## F2 score metric extension
f2_score <- MLMetric(
  function(observed, predicted, ...) {
    f_score(observed, predicted, beta = 2, ...)
  },
  name = "f2_score",
  label = "F2 Score",
  maximize = TRUE
)
```


## Usage

Once created, model and metric extensions can be used with the package-supplied fitting and performance functions.

```{r using_extensions_usage}
## Logistic regression analysis
library(MASS)
res <- resample(type ~ ., data = Pima.tr, model = LogisticModel)
summary(performance(res, metric = f2_score))
```
